Heuristics and Really Hard Instances for Subgraph Isomorphism Problems

### 2016-01-31 paper/paper.tex
Title consistency
More stuff

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from a given
pattern graph to a given target graph which preserves adjacency---in essence, we are ``finding a
copy of'' the pattern inside the target. The \emph{induced} variant of the problem additionally
requires that the mapping preserve non-adjacency, so there are no ``extra edges'' in the copy of the
pattern that we find. We illustrate both variants in \cref{figure:sip}.
Despite these problems being \NP-complete, modern practical subgraph isomorphism algorithms can
handle problem instances with many hundreds of vertices in the pattern graph, and up to ten thousand
vertices in the target graph \citep{Cordella:2004,Solnon:2010,Audemard:2014,McCreesh:2015}, leading
to successful application in areas such as computer
vision \citep{Damiand:2011,Solnon:2015}, biochemistry \citep{Giugno:2013}, and pattern recognition
\citep{Conte:2004}.

@@

However, these algorithms cannot handle \emph{arbitrary} instances of this size. The experimental
evaluations of these algorithms were performed using a mix of real-world graphs, graphs that encode
biochemistry and computer vision problems, and randomly generated graph pairs. Using random
instances to evaluate algorithm behaviour can be beneficial, because it provides a way of generating
many instances cheaply, and reduces the risk of over-fitting when tuning design parameters. The
random instances used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010},
which were generated by taking a random subgraph of a random graph (using various models, including
Erd\H{o}s-R\'enyi, scale-free, bounded degree, and meshes) and permuting the vertices. Such
instances are guaranteed to be satisfiable---\citet{Anton:2009} exploited this property to create
large sets of random satisfiable boolean satisfiability instances.  However, since this has been the
only approach used to generate random subgraph isomorphism instances, existing benchmark suites
contain relatively few non-trivial unsatisfiable instances, and the satisfiable instances tend to be
computationally fairly easy, with most of the difficulty being in dealing with the size of the
model.  This problem cannot be addressed by using a pattern graph from one of the random suites with
the ``wrong'' target graph: this tends to give either a trivially unsatisfiable instance, or a
satisfiable instance. (In particular, it is \emph{not} the case that a relatively small random graph
is unlikely to appear in a larger random graph.)

@@

\section{Induced Isomorphisms}

\begin{figure*}[p]
    \input{gen-graph-induced.tex}
    \caption{Behaviour of algorithms on the induced variant, shown in the style
    of \cref{figure:non-induced}. The second, third and fourth rows show the number of search nodes used by the
    Glasgow, LAD and VF2 algorithms. The fifth row shows a bound on the satisfiable region, by
    considering where a \emph{non-}induced isomorphism may also be a non-induced isomorphism between
    complement graphs. The sixth row plots constrainedness: the darkest region is where
    constrainedness is exactly one, and the lighter regions show where the problem is either over-
    or under-constrained. The final row shows when the Glasgow algorithm performs better when given
    the complements of the pattern and target graphs as
inputs.}\label{figure:induced}
\end{figure*}

@@

For small patterns, simply considering the mirror image of the non-induced prediction provides a
reasonable approximation of the location of the phase transition, but as we show on the fifth row of
the figure, this breaks down for larger patterns. For a better prediction, we repeat the argument
for locating the non-induced phase transition and additionally considering non-edges, to get an
expected number of solutions of \[ \langle Sol \rangle = t^{\underline{p}} \cdot {d_t}^{d_p \cdot
\binom{p}{2}} \cdot {(1 - d_{t})}^{(1 - d_{p}) \cdot \binom{p}{2}} \textnormal{.} \] We plot
this using black lines on \cref{figure:induced}---again, our prediction is accurate except for
very sparse or very dense patterns.

@@

?? Explain bottom row of \cref{figure:induced}, and give way of calculating it. The key point is
that depending upon where we are, changes to one of the two terms has more effect in the equation.
Note how sharp and well-defined this is.

?? Note that when the pattern is small, it is always best to try to move towards the satisfiable
region, even for unsatisfiable instances. It's not clear what's going on when the pattern is large.
Note the suggestion by \citet{Walsh:1998} that switching heuristics based upon an estimate of the
solubility of the problem may offer good performance.

?? Is this heuristic better in general, on non-random instances? Unfortunately most of the instances
in the standard benchmark set are too sparse for it to kick in. However, it provides a justification
for reusing the non-induced heuristics.

@@

?? Explain the flip bound on row 5, and note that the difficult region is roughly near the phase
transitions in the original graphs.

?? Note that the large unexpectedly hard region in the middle is solver-independent. Perhaps
subgraph isomorphism algorithms are not able to make use of adjacency and non-adjacency
simultaneously, and this is a weakness in how all three are designed? In the following section we
debunk this wonderful hypothesis.

?? Constrainedness predicts the hard region in the middle \citep{Gent:1996:Kappa}! Add a two line
description of kappa, the slightly modified way we calculated it. We plot this on the sixth row of
\cref{figure:induced}.

?? Note that the size of the state space is defined differently to in binary CSPs: we're using the
all diff constraint to determine the size of the state space, and are not treating it as a
constraint. In other words, our generate-and-test approach is alldiff-aware. This matters for kappa.

?? Maybe note a few solver-dependent hard instances.  There's odd unsat with LAD at the top right
(not EHPs), trivial for Glasgow: mapping a non-clique into a clique. Glasgow looks at non-edges and
can see immediately that no mapping is possible, but LAD doesn't do degree filtering on complements.

@@

The region in the parameter space where both pattern and target have medium density is far from a
phase transition, but nevertheless contains instances that are hard for current subgraph isomorphism
solvers. We would like to know whether this is due to a weakness in current solvers, or whether
instances in this region are inherently difficult to solve, as constrainedness suggests. In
\cref{figure:alt} we repeat the induced experiments on smaller pattern and target graphs, using
different solving techniques.  Although these techniques are not competitive in absolute terms, we
wish to see if the same pattern of behaviour occurs.

We create a pseudo-boolean (PB) encoding as follows. For each pattern vertex $v$ and each target
vertex $w$, we have a binary variable which takes the value 1 if and only if that $v$ is mapped to
$w$.  Constraints are added to ensure that each pattern vertex maps to exactly one target vertex,
that each target vertex is mapped to by at most one pattern vertex, that adjacent vertices are
mapped to adjacent vertices, and that non-adjacent vertices are mapped to non-adjacent vertices. We
used the Clasp solver~\citep{gekakaosscsc11a} version 3.1.3 to solve the pseudo-boolean instances.
The instances that are hard for the Glasgow solver remain hard for the PB solver, including
instances inside the central region, and the easy satisfiable instances remain easy.

We implemented an equivalent SAT encoding with direct encoding of the cardinality constraints. Using
the Glucose solver, we saw very similar performance to the PB encoding. We also implemented a
similar integer program encoding; the Gurobi solver was only able to solve some of the trivial
satisfiable instances, and was almost never able to prove unsatisfiability within the time limit. We
do not plot the results for either of these solvers.

@@

These results suggest that the central region may be genuinely hard, despite not being near a phase
transition, and that constrainedness is the better measure of difficulty. The clique encoding in
particular rules out the hypothesis that subgraph isomorphism solvers only find this region hard due
to not reasoning simultaneously about adjacency and non-adjacency, since the association graph
encoding constraints consider compatibility rather than adjacency and non-adjacency.

?? Lends weight to kappa

@@

?? \citep{Lipets:2009} inexact solutions?

>>>>

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from a given
pattern graph to a given target graph which preserves adjacency---in essence, we are ``finding a
copy of'' the pattern inside the target. The \emph{induced} variant of the problem additionally
requires that the mapping preserve non-adjacency, so there are no ``extra edges'' in the copy of the
pattern that we find. We illustrate both variants in \cref{figure:sip}.
Despite these problems being \NP-complete, modern practical subgraph isomorphism algorithms can
handle problem instances with many hundreds of vertices in the pattern graph, and up to ten thousand
vertices in the target graph \citep{Cordella:2004,Solnon:2010,Audemard:2014,McCreesh:2015}, leading
to successful application in areas such as computer
vision \citep{Damiand:2011,Solnon:2015}, biochemistry \citep{Giugno:2013,Carletti:2015}, and pattern
recognition \citep{Conte:2004}.

@@

However, these algorithms cannot handle \emph{arbitrary} instances of this size. The experimental
evaluations of these algorithms were performed using a mix of real-world graphs, graphs that encode
biochemistry and computer vision problems, and randomly generated graph pairs. Using random
instances to evaluate algorithm behaviour can be beneficial, because it provides a way of generating
many instances cheaply, and reduces the risk of over-fitting when tuning design parameters. The
random instances used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010},
which were generated by taking a random subgraph of a random graph (using various models, including
Erd\H{o}s-R\'enyi, scale-free, bounded degree, and meshes) and permuting the vertices. Such
instances are guaranteed to be satisfiable---\citet{Anton:2009} exploited this property to create
large sets of random satisfiable boolean satisfiability instances.  However, since this has been the
only approach used to generate random subgraph isomorphism instances, existing benchmark suites
contain relatively few non-trivial unsatisfiable instances, and the satisfiable instances tend to be
computationally fairly easy, with most of the difficulty being in dealing with the size of the
model.  ?? \citet{Battiti:2007}

This problem cannot be addressed by using a pattern graph from one of the random suites with the
``wrong'' target graph: this tends to give either a trivially unsatisfiable instance, or a
satisfiable instance. (In particular, it is \emph{not} the case that a relatively small random graph
is unlikely to appear in a larger random graph.)

@@

\section{Induced Subgraph Isomorphisms}

\begin{figure*}[p]
    \input{gen-graph-induced.tex}
    \caption{Behaviour of algorithms on the induced variant, shown in the style
    of \cref{figure:non-induced}. The second, third and fourth rows show the number of search nodes used by the
    Glasgow, LAD and VF2 algorithms. The fifth row shows a bound on the satisfiable region, by
    considering where a \emph{non-}induced isomorphism may also be a non-induced isomorphism between
    complement graphs. The sixth row plots constrainedness: the darkest region is where
    constrainedness is exactly one, and the lighter regions show where the problem is either over-
    or under-constrained. The final row shows when the Glasgow algorithm performs better when given
    the complements of the pattern and target graphs as inputs---the solid lines show the location
of the phase transition, and the dotted lines are $t_d=0.5$ and the $p_d=t_d$ diagonal.}\label{figure:induced}
\end{figure*}

@@

For small patterns, simply considering the $p_d=t_d$ diagonal mirror image of the non-induced
prediction provides a reasonable approximation of the location of the phase transition (recall the
complement property), but as we show on the fifth row of the figure, this breaks down for larger
patterns. For a better prediction, we repeat the argument for locating the non-induced phase
transition and additionally considering non-edges, to get an expected number of solutions of \[
\langle Sol \rangle = t^{\underline{p}} \cdot {d_t}^{d_p \cdot \binom{p}{2}} \cdot {(1 -
d_{t})}^{(1 - d_{p}) \cdot \binom{p}{2}} \textnormal{.} \] We plot this using black lines on the top
row of \cref{figure:induced}---again, our prediction is accurate except for very sparse or very
dense patterns.

@@

For patterns of order 10, it is always better to try to move towards the satisfiable region: if we
are in the bottom right diagonal half, we are best retaining the original heuristics (which move us
towards the top right), and if we are in the top left we should use the complement instead. This
goes against a suggestion by \citet{Walsh:1998} that switching heuristics based upon an estimate of
the solubility of the problem may offer good performance.

For larger patterns, more complex behaviour emerges. If we are the intersection of the bottom half
and the bottom right diagonal of the search space, we should always retain the original heuristic,
and if we are in the intersection of the top half and the top left diagonal, we should always use
the complements. This behaviour can be predicted by taking the partial derivatives of $\langle Sol
\rangle$ in the $-p_d$ and $t_d$ directions.  However, when inside the remaining two eights of the
parameter space, the partial derivatives of $\langle Sol \rangle$ disagree on which heuristic to
use, and using directional derivatives is not enough to resolve the problem. A close observation of
the data suggests that the actual location of the phase transition may be involved (and perhaps
\citeauthor{Walsh:1998}'s suggestion applies only in these conditions). In any case, $\langle Sol
\rangle$ is insufficient to explain the observed behaviour.

In practice, this is unlikely to be a problem: most real-world instances are extremely sparse and
are usually easy, which perhaps explains the continuing popularity of VF2's connectivity-based
heuristics \citep{Carletti:2015}. In this situation, these experiments justify re-using the
non-induced heuristics on induced problems.

@@

The region in the parameter space where both pattern and target have medium density is far from a
phase transition, but nevertheless contains instances that are hard for all three of our subgraph
isomorphism solvers. We would like to know whether this is due to a weakness in current solvers
(perhaps our solvers cannot region about adjacency and non-adjacency simultaneously?), or whether
instances in this region are inherently difficult to solve.  Thus we repeat the induced experiments
on smaller pattern and target graphs, using different solving techniques.  Although these techniques
are not competitive in absolute terms, we wish to see if the same pattern of behaviour occurs. The
results are plotted in \cref{figure:alt}.

@@

The pseudo-boolean (PB) encoding is as follows. For each pattern vertex $v$ and each target
vertex $w$, we have a binary variable which takes the value 1 if and only if that $v$ is mapped to
$w$.  Constraints are added to ensure that each pattern vertex maps to exactly one target vertex,
that each target vertex is mapped to by at most one pattern vertex, that adjacent vertices are
mapped to adjacent vertices, and that non-adjacent vertices are mapped to non-adjacent vertices. We
used the Clasp solver~\citep{gekakaosscsc11a} version 3.1.3 to solve the pseudo-boolean instances.
The instances that are hard for the Glasgow solver remain hard for the PB solver, including
instances inside the central region, and the easy satisfiable instances remain easy.

We also implemented an equivalent SAT encoding with direct encoding of the cardinality constraints.
Using the Glucose solver, we saw very similar performance to the PB encoding. We also implemented a
similar integer program encoding; the Gurobi solver was only able to solve some of the trivial
satisfiable instances, and was almost never able to prove unsatisfiability within the time limit. We
do not plot the results for either of these solvers.

@@

These results suggest that the central region may be genuinely hard, despite not being near a phase
transition. The clique encoding in particular rules out the hypothesis that subgraph isomorphism
solvers only find this region hard due to not reasoning simultaneously about adjacency and
non-adjacency, since the association graph encoding constraints consider compatibility rather than
adjacency and non-adjacency.

Constrainedness, denoted $\kappa$, is an alternative measure of difficulty designed to generalise
hardness parameters across different combinatorial problems \citep{Gent:1996:Kappa}. A problem with
$\kappa < 1$ is said to be underconstrained, and is likely to be satisfiable; a problem with $\kappa
> 1$ is overconstrained, and is likely to be unsatisfiable. Empirically, problems with $\kappa$
close to 1 are hard, and problems where $\kappa$ is very small or very large are usually easy. By
handling injectivity as a restriction on the size of the state space rather than as a constraint, we
derive \[ \kappa = 1 - \frac{\log t^{\underline{p}} \cdot {d_t}^{d_p \cdot \binom{p}{2}} \cdot {(1 -
d_{t})}^{(1 - d_{p}) \cdot \binom{p}{2}}}{\log t^{\underline{p}}}\] for induced isomorphisms, which
we plot on the sixth row of \cref{figure:induced}. We see that constrainedness predicts that the
central region will still be relatively difficult for larger pattern graphs: although the problem is
overconstrained, it is less overconstrained than in the regions the Glasgow and LAD solvers found easy.
Thus it seems that rather than just being a unification of existing generalised heuristic
techniques, constrainedness also gives a better predictor of difficulty than proximity to a phase
transition.  Unfortunately, constrainedness does not help us with heuristics: minimising
constrainedness gives the same predictions as maximising the expected number of solutions.

@@
